Stage-Splitting a Modal Language

### 2014-06-06 paper/paper.tex
starting metatheory notes
added more to semantics discussion

\usepackage{amsmath}
%\usepackage{amssymb}
\usepackage{bbm}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{mathpartir}
\usepackage{mathabx}
\usepackage{listings}

@@

\section{Diagonal Semantics}

@@

\begin{enumerate}
\item In absence of $\next$ expressions, it should be equivalent to standard call-by-value evaluation.
\item $\next$ and $\prev$ should be inverses of the other.
\item All of the first stage work should be completed before second stage work.
\item The result for a stage-2 expression should be the same as if all of the $\next$s and $\prev$s were interpreted as the identity function.
\end{enumerate}

@@

\begin{center}
\begin{tabular}{|c|p{2.5cm}|p{2.5cm}|} \hline
Judgment & Reads & Conditions \\ \hline 
$\isvalone e$ & ``$e$ is a $\bbone$-value under context $\Xi$'' & ... \\ \hline 
$\Xi \vdash \diaonesub$ & ``$e$ evaluates to future table $\gamma$ and value $v$'' 
& $\typesone e A$ 
	\newline $\isvalone v$ 
	\newline ... \\ \hline 
$\diatwosub$ & ``$e$ speculates to $q$'' & ... \\ \hline 
...&...&... \\ \hline
\end{tabular}
\end{center}

@@

\subsection{Adherence to Goals}

We meet our first (discuss) and second goals (show examples).

That we meet our third goal is clear from just the structure of our judgments (...).

Under the strictest interpretation, we do not meet our third goal.  Consider the following examples (show speculating down an if).

But as it turns out, a semantics which had the erasure behavior would not meet goal 3! (explain using the same example)

\subsection{Comparison to Other Semantics}

The differences between our diagonal semantics, the erasure semantics, and Davies' semantics can be characterized entirely by how they behave at $\next$ expressions:
\begin{itemize}
\item {\bf Erasure Semantics.} Evaluate contents of $\next$ immediately.
\item {\bf Diagonal Semantics.} Speculate to find the $\prev$s, evaluate them in place, then place the contents of $\next$ into a table and create a hatted variable to act as a value.
\item {\bf Davies' Semantics.} Speculate to find the $\prev$s, evaluate them in place, then treat the reduced $\next$ as a value.
\end{itemize}

>>>>

\usepackage{amsmath}
%\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{bbm}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{mathpartir}
\usepackage{mathabx}
\usepackage{listings}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}

@@

\section{Semantics}

There's more than one way to define a semantics for our language.  We start this section by considering two proposals, characterizing them mainly by how they handle values of $\fut$ type.

\subsection{The Erasure Semantics}

The first proposal we call the {\em erasure semantics}, because it corresponds to what one would get by simply {\em erasing} all of the $\next$ and $\prev$ terms that appear in \lamStaged.  In particular, we say that $\next v$ is a first-stage value of type $\fut A$ if and only if $v$ is a second-stage value of type $A$, (...)

We say that erasure semantics gives us a notion of multi-stage evaluation.  One judgement is multistage evaluation at 1, and the other is multistage evaluation at 2.  The downside of this is that the erasure semantics doesn't seem to do all of the stage-1 work before the stage-2 work, for some notion of ``before''.  Of course, the erasure semantics might be equivalent to some other semantics which does have that property.  Unfortunately, this is false (show second-stage if example).

The benefit of the erasure semantics is that it's very natural, and has a familiar cost model.  It also {\em obviously} produces the correct answer, so we can use the erasure semantics as a reference to prove the reasonableness of any other semantics.

\subsection{Meta Semantics}

A different semantics was provided in \cite{davies96}.  We briefly review a two-stage version of that semantics here.

In the erasure semantics, $\next v$ is a value only if $v$ is fully reduced.  But in the meta semantics, $\next e$ is a value only if $e$ has no $\prev$ terms; $v$ is allowed to have unreduced stage-2 computation.

The Davies semantics is comprised of two mutually recursive judgements: $\daviesz$ and $\davieso$.  For some $\colone e A$, the $\daviesz$ judgment evaluates only the first-stage parts of $e$, leaving unevaluated second-stage code within $\next$s.  (... This essentially gives us partial evaluation, and we're left to just evaluate the residual normally to get multi-stage evaluation...)

The benefit of the meta semantics is that it gives us a very explicit notion of partial evaluation.  This also, by construction, means that the meta semantics does all of the first stage-work {\em before} the second stage work begins.

The $\next$-by-name behavior of the meta semantics of course means that it happily duplicates second-stage code, which could increase the cost.  This makes reasoning about second-stage cost rather difficult.

\subsection{Our Semantics}

@@

\begin{enumerate}
\item Modulo termination, it should be equivalent to the erasure semantics.
\item All of the first stage work should be completed before second stage work.  Ideally, it should just have a notion of partial evaluation, like the meta semantics.
\item Should be $\next$-by-value, rather than $\next$-by-name, like the erasure semantics.  
\end{enumerate}

We meet all of these goals.

@@

\begin{center}
\begin{tabular}{|c|p{2.5cm}|p{2.5cm}|} \hline
Judgment & Reads & Conditions \\ \hline 
$\isvalone e$ & ``$e$ is a $\bbone$-value under context $\Xi$'' & ... \\ \hline 
$\Xi \vdash \diaonesub$ & ``$e$ evaluates to future table $\xi$ and value $v$'' 
& $\typesone e A$ 
	\newline $\isvalone v$ 
	\newline ... \\ \hline 
$\diatwosub$ & ``$e$ speculates to $q$'' & ... \\ \hline 
...&...&... \\ \hline
\end{tabular}
\end{center}

@@

\subsection{Other nice properties}

\begin{enumerate}
\item For some $\colone{e}{A}$ containing no $\next$ expressions, the semantics is derivationally equivalent to standard call-by-value evaluation.
\end{enumerate}

\subsection{Metatheory}

\TODO unfinished; where should this go? combine with table above

\noindent\TODO terminology for $\Xi,\Gamma,\xi,v,q$?

\noindent\TODO need to define 1-vals at 1 and 2, and 2-vals at 2

\noindent\TODO put $\Xi$ and $\Gamma$ in the judgments where appropriate

% local macros; put them in macros.tex once they are more finalized
\newcommand{\wf}{\ \mathsf{wf}}
\newcommand{\dom}[1]{\mathsf{dom}(#1)}
\newcommand{\valoo}{\ \bbone\textsf{-}\mathsf{val} @ \bbone}
\newcommand{\valot}{\ \bbone\textsf{-}\mathsf{val} @ \bbtwo}
\newcommand{\valtt}{\ \bbtwo\textsf{-}\mathsf{val} @ \bbtwo}

\begin{definition}
Contexts $\Xi$ and $\Gamma$ are well-formed ($\Xi\wf$, $\Gamma\wf$) if they
contain only stage-2 variables.
\end{definition}

\begin{theorem}
If $\Xi;\typesone e A$ then $\Xi\wf$, $\Gamma\wf$, and $A\istypeone$.
\end{theorem}

\begin{theorem}
If $\Xi;\Gamma\vdash\diaonesub$ then:
\begin{enumerate}
\item $\typesone e A$ for some 1-type $A$;
\item $v\valoo$;
\item \TODO unfinished
\end{enumerate}
\end{theorem}
